### 2017-01-22 doc/manuscript.Rmd
First (rough) draft of discussion - prev lit and interpret

TAG as a pharmaceutical target and as a measure of metabolic health is a single
measure of all fatty acids within the TAG fraction. However, given the 
increasing appreciation for the role of the specific fatty acid (FA) composition
of various blood lipid fractions on health outcomes, reducing the complex FA 
metabolism within the TAG fraction to a single measure may contribute to the 
less than expected {{?}} changes in risk reduction.

@@

The PLS results for the insulin sensitivity (`r cite_f('fig_pls')`) revealed a clustering 
of the FA 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
the FA fell between the dashed and solid circles) and negatively on HOMA2-%S
and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3, and 22:6n-3 loaded positively
on both insulin sensitivity measures. No other TAGFA loaded strongly, the
results of which remained after running the PLS model on the cross-validated
testing set. The PLS model for insulin sensitivity had good predictive ability, 
determined by correlating the predicted outcome values with the actual values
(HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

@@

Previous research has shown that carbohydrate intake increases DNL {{ cite }}. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for diabetes {{ cite }}
as well as studies showing higher refined carbohydrates (e.g. added sugars) and
a higher risk for diabetes {{cite}}. Our study extends {{adds to? confirms?}}
these findings on DNL by showing that TAGFA with 14 to 16 carbons all strongly
predicted lower insulin sensitivity. While these FA also had a significant
association with beta-cell function, the results were not as strong as with the
IR. {{ why is it that DNL increases risk? }}

To our knowledge, no longitudinal study has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using 
statistical techniques designed to adequately analyze ... {{shots fired? 
necessary?}}.

To our knowledge, two prospective studies have been published that had similar
research objectives as the present analysis. One was a nested case-control
cohort (n=189 cases and n=189 controls) of the Framingham offspring study
[@Rhee2015a], which found that TAGFA composition with a lower carbon chain and less double
bonds (e.g. 14:0, 16:0) had a higher OR {{confirm}} for diabetes after 12-years
while higher carbon chain and more double bond TAGFA had a lower OR for
diabetes. This pattern of TAGFA with diabetes risk remained
the same when they examined HOMA-IR cross-sectionally at the baseline visit. The
other study was a prospective cohort of males in Finland [@Lankinen2015a], of
which they had TAGFA data on 831 participants after 6-years from the baseline
visit. This cohort has similar outcomes as PROMISE, however, OGTT data was only
available at the 6-year visit. A beta-cell function measure was computed, but
only included calculations up to 30 minutes of the OGTT. As with our results,
they found a larger effect size for insulin sensitivity compared to the
beta-cell function. The specific TAGFA that associated with the metabolic
measures were similar between studies, except they did not find {{confirm}}

In addition, the Framingham offspring study [@Rhee2011a] did not have data on
beta-cell function while the Finnish cohort [@Lankinen2015a] used {{ how do I
say this politely Tony? They're methods and article was ... questionable... }}
simple statistical techniques to analysis the longitudinal data, which do not
adequately take into consideration auto-correlation of longitudinal values and
the high dimensionality of the TAGFA composition data structure.

A smaller study (n=16) of mostly females using highly detailed lipodomic
techniques found a positive correlation between total esterified (of which TAGFA
make up the majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR [@Kotronen2009a].

Limitations with estimated DNL or measured DNL in prev studies.

Several studies have been conducted that link greater hepatic DNL with the 14
and 16 carbon FA. A short, small feeding trial (n=24) was conducted to identify
which biomarkers most accurately reflected DNL [@Lee2015a]. The study found that
16:1n-7, directly measured DNL using isotopes, and liver fat were all highly
correlated with each other, suggesting that 16:1n-7 is a good biomarker for
hepatic DNL.
In another small (n=14) feeding trial, meal type (high fat vs low fat) was
tested to determine the effect on DNL and TAGFA composition [@Wilke2009a]. They
found that 14:0, 16:0, 16:1, and 18:2 were higher in the low fat (high
carbohydrate) group. These fatty acids are similar to the fatty acids we found
that clustered together, suggesting {{another word}} that these fatty acids are
indicative of a higher carbohydrate diet.
Suggesting {{expand/synth}} that greater hepatic fat may also contribute to greater DNL,
increasing the 14 and 16 carbon FA in circulation, potentially exposing tissues
to greater lipotoxicity, resulting in further hepatic dysfunction and continuing
the cycle.

@@

- A weight loss intervention trial showed that changes in the TAGFA composition
of 19 participants accompanied changes in the weight and insulin sensitivity over
the 33 week period [@Schwab2008a]. Specifically, the TAGFA 14:0, 14:1, 16:0,
16:1, 18:0, 18:3n-3, and 18:3n-6 were all lower after weight loss and correlated
with improved insulin sensitivity.
{{ hard to disentangle reverse causality here }}
    - Schwab2008a (TAGFA composition and insulin sensitivity) (n=75 (but only
    n=19 for lipidomics), overweight or obese, weght loss intervention (33
    weeks), weight, FSIGT, GC, UPLC-MS, after weight loss 14, 141, 16, 161, 18,
    183n3, 183n6 were lower, lower TAGFA that contained 14, 14:1, 16, 161, 18,
    181 were correlated with improved Si)

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL {{cite}}. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
in adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
VLDL-TAG during fasting and ~40% during fed while dietary fatty acids
contributed ~25% to newly synth VLDL-TAG {{fasting or fed?}} [@Barrows2006a].
While the NEFA fraction is much smaller than the TAG fraction even though it
contributes substantially to the TAG fraction, VLDL-TAG is not cleared
(lipolyzed) as much as chylomicron (which also contains a large amount of TAG)
and so has a higher half-life than NEFA, remaining in circulation longer {{hence
why there is bigger TAG pool vs NEFA}}. In an analysis of the NEFA fraction in
the PROMISE cohort with the OGTT-derived outcomes (currently unpublished)
{{correct word for this??}}, we found that total NEFA, but not the specific
composition associated with beta-cell function and not with insulin resistance.
While NEFA may contribute substantially to the TAG fraction, there may be
something unique about the specific composition of the TAGFA fraction that may
allow it to enter insulin sensitive tissues more easily than NEFA. It may be
that VLDL receptors allow FA to transition to these tissue compared to NEFA
bound to albumin {{ confirm and cite }}, and thus able to exert potential
lipotoxic effects. {{ comment on comparing models with and without NEFA }}

NEFA may be a mediating factor in association, but that adjustment will not
attenuate this assoc since TAGFA composition is also dictated by other sources
for DNL (e.g. recent meal, DNL from glucose stores, etc.).

As with our previous analysis of the PL fatty acid fraction {{cite}}, we found
that cis-vaccenic acid strongly predicted higher insulin sensitivity and
beta-cell function. Likewise, in the TAGFA fraction we found a similar
association with cis-vaccenic acid. {{ continue? }}
    - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
    two groups: normogly and hypergly. results in general showed no diff between
    groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
    not present results on whole fraction for IR)

{{ comment on unadjusted vs adjusted }}

{{ commment on null interactions with time }}
- Challenge longitudinally since muscle IR can promote DNL + TAG Flannery2012a
Since no interaction, no loop? Not enough sample, already IR? maybe?

@@

Prospective cohorts:
    - Chien K, et al. A prediction model for type 2 diabetes risk among Chinese people. Diabetologia. 2009;52(3):443–450.
    - D’Agostino RB Jr, Hamman RF, Karter AJ, Mykkanen L, Wagenknecht LE, Haffner SM. Cardiovascular disease risk factors predict the development of type 2 diabetes: the insulin resistance atherosclerosis study. Diabetes Care. 2004;27(9):2234–2240.
    - Schulze MB, et al. Use of multiple metabolic and genetic markers to improve the prediction of type 2 diabetes: the EPIC-Potsdam Study. Diabetes Care. 2009;32(11):2116–2119.

>>>>

TAG as a pharmaceutical target and as a measure of metabolic health is a single
measure of all fatty acids within the TAG fraction. However, given the 
increasing appreciation for the role of the specific fatty acid composition
of various blood lipid fractions on health outcomes, reducing the complex fatty acids 
metabolism within the TAG fraction to a single measure may contribute to the 
less than expected {{?}} changes in risk reduction.

@@

The PLS results for the insulin sensitivity (`r cite_f('fig_pls')`) revealed a clustering 
of the fatty acids 14:0, 14:1n-7, 16:0, and 16:1n-7. These TAGFA loaded strongly (meaning
the fatty acids fell between the dashed and solid circles) and negatively on HOMA2-%S
and ISI, suggesting this cluster of TAGFA tracks together with lower insulin
sensitivity. The TAGFA 20:2n-6, 20:5n-3, 22:5n-3, and 22:6n-3 loaded positively
on both insulin sensitivity measures. No other TAGFA loaded strongly, the
results of which remained after running the PLS model on the cross-validated
testing set. The PLS model for insulin sensitivity had good predictive ability, 
determined by correlating the predicted outcome values with the actual values
(HOMA2-%S: `r cor_pls_homa2$r_p`; ISI: `r cor_pls_isi$r_p`).

@@

To our knowledge, no longitudinal study has examined the role of the composition
of the TAGFA fraction on detailed OGTT-derived metabolic measures using 
advanced {{another word?}} statistical techniques designed to maximally utilize
the data. Two large prospective studies have been published that had similar
research objectives as the present analysis. One was a nested case-control
cohort (n=189 cases and n=189 controls) of the Framingham offspring study
[@Rhee2015a], which found that TAGFA composition with a lower carbon chain and
less double bonds (e.g. 14:0, 16:0) had a higher risk for diabetes after 
12-years while higher carbon chain and more double bond TAGFA had a lower risk 
for diabetes. This pattern of TAGFA with diabetes risk remained the same when 
they examined HOMA-IR cross-sectionally at the baseline visit. The other study 
was a prospective cohort of males in Finland [@Lankinen2015a], of which they had
TAGFA data on 831 participants after 6-years from the baseline visit. This 
cohort has similar outcomes as PROMISE, however, OGTT data was only available at
the 6-year visit. A beta-cell function measure was computed, but only included 
calculations up to 30 minutes of the OGTT. As with our results, they found a 
larger effect size for insulin sensitivity compared to the beta-cell function. 
The specific TAGFA that associated with the metabolic measures were similar 
between studies, particularly with the four DNL fatty acids. A limitation of 
this study was the simpler statistical techniques used to analyze the data, 
which is not able to adequately use the longitudinal and high-dimensional 
dataset. {{ They're article and methods were ... questionable... }} In another 
study of a much smaller (n=16) mostly female group, the authors report a 
positive correlation between total esterified (of which TAG make up the 
majority) 16:0, 16:1n-7, and 18:1n-9 with HOMA-IR [@Kotronen2009a], similar as 
our results except for 18:1n-9.

Previous research has shown that carbohydrate intake increases DNL 
[@Hodson2008a; @Kawano2013a; @Harding2015a; @Hudgins2000a; @Parks1999a]. In
particular, DNL from refined or simple carbohydrate sources increases the 14 to
16 chain fatty acids as well as the 18 chain TAGFA. Several studies have shown a
link between higher estimated DNL and an increased risk for metabolic 
dysfunction [@Ma2015a; @Zong2013a; @Lankinen2015a; @Kroger2011a] as well as
studies showing higher refined carbohydrates (e.g. added sugars) and a higher
risk for diabetes {{cite}}. Our study extends these findings on DNL by showing
that TAGFA with 14 to 16 carbons all strongly predicted lower insulin
sensitivity. While these fatty acids also had a significant association with
beta-cell function, the results were not as strong as with the insulin
sensitivity.

The link between higher DNL and increases in specific fatty acids has been
examined in several studies. Previous studies have examined used markers
of DNL, such as the ratio between 18:2n-6 to 16:0 or 16:0 to 14:0 ratio
[@Hodson2008a; @Lankinen2015a; @Kroger2011a]. However, there are major
limitations to using these ratios as the fatty acids used in their calculation
can be created from DNL, they are also obtained from the diet [@Hodson2008a]. A
short, small feeding trial (n=24) was conducted to identify fatty acids that
most accurately reflected DNL, as potential biomarkers [@Lee2015a]. The study
found that palmitoleic acid (16:1n-7), directly measured DNL using isotopes, and
liver fat were all highly correlated with each other (r>0.50), suggesting that
16:1n-7 may be a good biomarker for hepatic DNL. In another small (n=14) feeding
trial, meal type (high fat vs low fat) was tested to determine its effect on DNL
and TAGFA composition [@Wilke2009a]. The authors reported that 14:0, 16:0, 16:1,
and 18:2 were higher in the low fat (high carbohydrate) group. These fatty acids
are similar to the fatty acids we found that clustered together, implicating 
these fatty acids as indicative of a higher carbohydrate diet. A higher
carbohydrate diet, particularly with simple carbohydrates, may lead to greater
DNL in an attempt to control blood glucose, increasing hepatic fat stores, and
increasing the amount of TAGFA in circulation. The increase in circulating 14
and 16 carbon fatty acids may then increase the exposure of tissues to lipotoxicity,
for instance from palmitic acid (16:0), which is well-known to have harmful
effects on tissues [@Riserus2008a; @Iggman2010a].

We found no interaction by time for any of the TAGFA on the metabolic outcomes.
Analyzing these biological processes longitudinally is challenging due to the
potential feedbacks involved. For instance, while greater DNL may promote muscle
resistance, the reverse may also be true [@Flannery2012a]. Higher insulin
resistance may encourage greater DNL to handle the higher blood glucose. To
illustrate, in a weight loss intervention trial (n=19 with TAGFA data),
participants who lost weight over 33 weeks showed higher insulin sensitivity and
lower TAGFA composition indicative of lower DNL (less 14:0, 14:1, 16:0, 16:1,
etc) [@Schwab2008a]. Given the complex biological mechanisms and feedback loops
involved, disentangling whether insulin sensitivity influences TAGFA more
strongly than TAGFA influencing insulin sensitivity will require complicated
research designs and analyses. Our own null results on this question may
tentatively point to an equal feedback mechansim between TAGFA and insulin
sensitivity.

Serum TAG are packaged into lipoproteins, particularly VLDL, by the liver. The
sources of fatty acids for TAG production during fasting are stored fatty acids
within the liver, circulating NEFA, and through DNL [@Barrows2006a]. Therefore, the
TAGFA composition reflects not just dietary intake but also stored fatty acids
from the adipose tissue and the DNL capacity of the individual. During fasting, a
large portion of fatty acids bound together as TAG are obtained from circulating
NEFA. A study examining this found that NEFA contributed 60-80% of newly synth
VLDL-TAG during fasting and ~40% during fed while dietary fatty acids
contributed ~25% to newly synth VLDL-TAG [@Barrows2006a]. In an analysis of the
serum NEFA fraction in the PROMISE cohort with the OGTT-derived outcomes
(currently unpublished), we found that higher total NEFA but not the specific
composition associated with lower beta-cell function and not with insulin
resistance. While NEFA may contribute substantially to the TAG fraction, there
may be something unique about the specific composition of the TAGFA fraction
that may allow it to enter insulin sensitive tissues more easily than NEFA. It
may be that VLDL receptors allow fatty acids to enter these tissue more easily than 
NEFA bound to albumin, or it may be due to the other influences on TAGFA
composition such as dietary intake.

{{ Not sure about this... Include this? }}
As with our previous analysis of the PL fatty acid fraction {{cite}}, we found
that cis-vaccenic acid strongly predicted higher insulin sensitivity and
beta-cell function. Likewise, in the TAGFA fraction we found a similar
association with cis-vaccenic acid. {{we can discuss to include more}}

<!--
{{ comment on comparing models with and without NEFA }} While
NEFA may be a mediating factor, we found that adjustment for it did not
{{confirm}} attenuate this association since TAGFA composition is also dictated
by other sources for DNL (e.g. recent meal, DNL from glucose stores, etc.).

While the NEFA fraction is much smaller than the TAG fraction, even though it
contributes substantially to the TAG fraction, the VLDL-TAG is not cleared
(lipolyzed) as much as chylomicron (which also contains a large amount of TAG)
and so has a higher half-life than NEFA, remaining in circulation longer {{hence
why there is bigger TAG pool vs NEFA}} {{cite?}}.

@@

    - Zulyniak2012a (only vaccenic acid in TAG) (n=20, TLC-GC, HOMA-IR/HOMA-B,
    two groups: normogly and hypergly. results in general showed no diff between
    groups, only cis-vaccenic was higher in hyper, 'trend' with palmitic, did
    not present results on whole fraction for IR)
{{ comment on unadjusted vs adjusted }}

{{ commment on null interactions with time }}

@@

Prospective cohorts:
    - Chien K, et al. A prediction model for type 2 diabetes risk among Chinese people. Diabetologia. 2009;52(3):443–450.
    - D’Agostino RB Jr, Hamman RF, Karter AJ, Mykkanen L, Wagenknecht LE, Haffner SM. Cardiovascular disease risk factors predict the development of type 2 diabetes: the insulin resistance atherosclerosis study. Diabetes Care. 2004;27(9):2234–2240.
    - Schulze MB, et al. Use of multiple metabolic and genetic markers to improve the prediction of type 2 diabetes: the EPIC-Potsdam Study. Diabetes Care. 2009;32(11):2116–2119.
-->
